---
title: "t and Permutation Test Homework"
author: "Megha Pandit"
date: "April 16, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Snodgrass Problem

```{r}
twain <- c(0.225, 0.262, 0.217, 0.240, 0.230, 0.229, 0.235, 0.217)
snodgrass <- c(0.209, 0.205, 0.196, 0.210, 0.202, 0.207, 0.224, 0.223, 0.220, 0.201)
```

The null hypothesis here is: $H_{0}: \mu_{t} = \mu_{s}$, and alternate hypothesis is: $H_{1}: \mu_{t} \neq \mu_{s}$

For the parametric test, the two test statistics we need to test for are:
$$T = \frac{\bar{X_t} - \bar{X_s}}{S_{pool}\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$
and 
$$S^{2}_{pool} = \frac{(n_1 - 1)S^2_1(n_2 - 1)S^2_2}{n_1 + n_2 + 2}$$

```{r}
spool <- sqrt(((8-1)*var(twain)+(10-1)*var(snodgrass))/(8+10-2))
t <- (mean(twain)-mean(snodgrass))/spool*sqrt((1/8)+(1/10))
```

The t statistic is 0.87, which is less than 2. Therefore, we cannot reject the null hypothesis.

For the permutation test:
Considering the large number of permutations we could get from the data, only 2000 are considered here.

```{r, fig.height=3, fig.width=4}
mean_diff <- mean(twain) - mean(snodgrass)
data <- c(twain, snodgrass)

set.seed(99)
p.test <- function(x, a, b){
s <- sample(x)
mean(s[1:a]) - mean(s[(a+1):b])
}
many <- replicate(2000, p.test(data,8,18))
hist(many, main = "Histogram of Difference", xlab = "Difference in Mean"); abline(v=mean_diff,lwd=2, col="purple")
```



```{r}
n <- sum(abs(many) > abs(mean_diff));n
```
There are only two out of the 2000 times where the true difference was exceeded and the p value is < 0.05, therefore, we can reject the null hypothesis. 

#Hot Dog Problem

```{r}
hd <- c(186, 181, 176, 149, 184, 190, 158, 139, 175, 148, 152, 111, 141, 153, 190, 157, 131, 149, 135, 132)
hd_bar <- mean(hd)
hd_bar #mean of the observations

hd_sd <- sd(hd)
hd_sd #standard deviation of the observations

n <- 20
```

The critical value of t is 1.729, for a two-tailed test at an $\alpha = 0.1$ and df = 19.

```{r}
#upper tail
hd_bar + 1.729 * hd_sd/sqrt(n)
```

```{r}
#lower tail
hd_bar - 1.729 * hd_sd/sqrt(n)
```

The 90% confidence interval for the mean is [148.0962, 165.6038].

#Reading Score Problem

```{r}
c1 <- c(1.23, 1.42, 1.41, 1.62, 1.55, 1.51, 1.60, 1.76)
c2 <- c(1.76, 1.41, 1.87, 1.49, 1.67, 1.81)
t.test(c1, c2, alternative= "less", var.equal = TRUE)
```

```{r, fig.height=3, fig.width=4}
m_diff <- abs(mean(c1) - mean(c2))
c <- c(c1, c2)
set.seed(222)
diff <- replicate(1000, p.test(c, 8, 14))
hist(diff, main = "Histogram of Difference", xlab = "Difference in Mean"); abline(v=m_diff,lwd=2, col="green")

num <- sum(abs(diff)>m_diff); num
```

In the t-test, p > 0.05, therefore, we fail to reject the null hypothesis that the mean of class 1 is higher or equal to the mean of class 2. 

In the permutation test, p = 0.11 > 0.05, and similar to the t-test, we fail to reject the null hypothesis. 

