---
title: "Homework 3"
author: "Megha Pandit"
date: "March 8, 2019"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Hmisc)
library(base)
```

#Problem 1

```{r}
x1 <- read.table("maybe_uniform.txt")
data1 <- c(x1[1,], x1[2,], x1[3,], x1[4,], x1[5,])
data1 <- as.numeric(data1)
ks.test(data1, "punif")

hist(data1, breaks = 10)
```
__**Given the p-value of 0.35 from the KS test, we cannot reject the null hypothesis that the data comes from a uniform distribution.**__


```{r, message=FALSE, warning=FALSE}
x2 <- seq(-20, 20, by = 0.01)
fx2 <- ifelse(0 < x2 & x2 <= 0.5, 3/2, ifelse(0.5 < x2 & x2 < 1, 1/2, 0))
ks.test(fx2, data1)
```
__**The resulting p-value suggests that we reject the null hypothesis that the above two samples come from the same distribution. The D statistic is much larger for this model than for the uniform distribution, implying that the above model is not better than uniform distribution.**__


```{r}
ks.test(data1, "pgamma", shape = 1, rate = 2)
```
__**The D statistic is smaller for the above test, and the p-value also suggests that we cannot reject the null hypothesis that the data is drawn from a gamma distribution. This may be a better model than the previous two.**__

$~$
$~$

#Problem 2
```{r}
x3 <- read.table("maybe_normal.txt")
data3 <- c(x3[1,], x3[2,], x3[3,], x3[4,], x3[5,], x3[6,], x3[7,], x3[8,], x3[9,], x3[10,])
data3 <- as.numeric(data3)

ks.test(data3, "pnorm", mean = 26, sd = 2)
```

```{r}
hist(data3)
qqnorm(data3)
qqline(data3)
```

__**From the above results, and with a p-value of 0.9663, we cannot reject the null hypothesis that the data comes from normal distribution.**__

$~$
$~$

#Problem 3
```{r, message=FALSE, warning=FALSE}
txt1 <- read.table("maybe_same_1.txt")
X <- c(txt1[1,], txt1[2,], txt1[3,], txt1[4,], txt1[5,])
X <- as.numeric(X)

txt2 <- read.table("maybe_same_2.txt")
Y <- c(txt2[1,], txt2[2,], txt2[3,], txt2[4,], txt2[5,])
Y <- as.numeric(Y)

ks.test(X, Y)

ks.test(X+2, Y)
```

__**From the two-sample KS test for X and Y, we cannot reject the null hypothesis that X and Y come from the same distribution. However, from the two-sample test for X+2 and Y, the p-value is 0.0001673 suggesting that we could reject the null hypothesis that X+2 and Y are from the same distribution.**__

$~$
$~$

#Problem 4

```{r}
norm1 <- readRDS("norm_sample.RData")
norm1 <- data.frame(norm1)
ns <- ecdf(norm1$norm1)

norm1$ecdf <- ns(norm1$norm1)
norm1$nm <- pnorm(norm1$norm1)
norm1$D <- norm1$ecdf - norm1$nm

print(paste0("The D statistic is: ", max(abs(norm1$D))))
```

```{r}
ks.test(norm1$norm1, "pnorm", mean = 0, sd = 1)
```

__**The calculated D statistic is 0.137 and the D statistic given by the KS test is 0.177. From the results of the KS test, we fail to reject the null hypothesis that the data comes fom a standard normal distribution.**__

$~$
$~$

#Problem 5

```{r}
fiji <- read.table("fijiquakes.dat", header = T)

fiji_ecdf <- ecdf(fiji$mag)

#Finding the 95% CI
ci_data <- sum(fiji$mag <= 4.9 & fiji$mag > 4.3)

print(paste0("The 95% confidence interval for F(4.9) - F(4.3): "))
binconf(ci_data, length(fiji$mag), method = "wilson", 0.05)
```

```{r}
faith <- read.table("faithful.dat", header = T, skip = 25)

faith_ecdf <- ecdf(faith$waiting)

#90% confidence interval for mean waiting time
mean_faith <- mean(faith$waiting)
sd_faith <- sd(faith$waiting)/sqrt(length(faith$waiting))

print(paste0("The 90% confidence interval for the mean waiting time is: [", mean_faith - 1.64*sd_faith,
             ", ", mean_faith + 1.64*sd_faith, "]"))

```

```{r}
summary(faith_ecdf)[3]

print(paste0("The estimated median waiting time is: ", summary(faith_ecdf)[3], " ", "min"))
```


